---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rirods

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/MartinSchobben/rirods/branch/master/graph/badge.svg)](https://app.codecov.io/gh/MartinSchobben/rirods?branch=master)
[![R-CMD-check](https://github.com/MartinSchobben/rirods/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/MartinSchobben/rirods/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The rirods package is an R client for iRODS.

## Installation

You can install the development version of rirods like so:

``` r
# install.packages("devtools")
devtools::install_github("MartinSchobben/rirods")
```

## Example

This is a basic example which shows you how to quickly launch a local iRODS server, connect with R, and perform some actions. For more information on the iRODS demo server check https://github.com/irods/irods_demo.

```{bash, eval=FALSE}
# clone the repo
# git clone https://github.com/irods/irods_demo
# initiate git submodules
# git submodule update --init
# to start
cd ../irods_demo
docker-compose up
```

```{r admin, include=FALSE}
library(rirods)
# switch to new irods project
create_irods("http://localhost/irods-rest/0.9.2", overwrite = TRUE)

# authenticate
iauth("rods", "rods")

# add user bobby
rirods:::iadmin(action = "add", target = "user", arg2 = "bobby", arg3 = "rodsuser")

# modify pass word bobby
rirods:::iadmin(action = "modify", target = "user", arg2 = "bobby", arg3 = "password", arg4  = "passWORD")
```

Provide the host name of the server to connect an R project with an iRODS server, like so:

```{r start, eval=FALSE}
library(rirods)

# connect project to server
create_irods("http://localhost/irods-rest/0.9.2")
```

In this example Bobby is a user of iRODS and he can authenticate himself with `iauth()`. This prompts a dialog where you can enter your username and password without hardcoding this information in your scripts.

```{r, bobby, eval=FALSE}
# login as bobby with password "passWORD"
iauth() # or iauth("bobby", "passWORD")
```


```{r, secret, include=FALSE}
# login as bobby
iauth("bobby", "passWORD")
```

Suppose Bobby would like to upload an R object from his current R session to an iRODS collection. For this, use the `iput()` command:

```{r put}
# some data
foo <- data.frame(x = c(1, 8, 9), y = c("x", "y", "z"))

# check where we are
ipwd()

# store
iput(foo)
```

To truly appreciate the strength of iRODS, we can add some metadata that describes the data object "foo". 

```{r meta}
# add some metadata
imeta(
  "foo", 
  "data_object", 
  operations = 
    list(operation = "add", attribute = "foo", value = "bar", units = "baz")
)

# check if file is stored with associated metadata
ils(metadata = TRUE)
```

If Bobby wanted to copy the foo R object from an iRODS collection to his local directory, he would use `iget()`:

```{r get}
# retrieve in native R format
iget("foo")
```

Possibly Bobby does not want a native R object to be stored on iRODS but a file type that can be accessed by other programs.

```{r filetypes}
library(readr)

# creates a csv file of foo
write_csv(foo, "foo.csv")

# send file
iput("foo.csv")

# check whether it is stored
ils()
```

```{r rmfile1, include= FALSE}
file.remove("foo.csv")
```

Later on somebody else might want to download this file again and store it locally.

```{r csv}
# retrieve it again later
iget("foo.csv")
read_csv("foo.csv")
```

By adding metadata you and others can more easily discover data in future projects. Objects can be searched with General Queries and `iquery()`:

```{r query, options}
# look for objects in the home directory with a wildcard `%`
iquery("SELECT COLL_NAME, DATA_NAME WHERE COLL_NAME LIKE '/tempZone/home/%'")
```

```{r query2}
# or where data objects named "foo" can be found
iquery("SELECT COLL_NAME, DATA_NAME WHERE DATA_NAME LIKE 'foo%'")
```

Finally, we can clean up Bobby's home directory:

```{r clean}
# delete object
irm("foo", trash = FALSE)
irm("foo.csv", trash = FALSE)

# check if object is removed
ils()
```


```{r rmfile2, include= FALSE}
file.remove("foo.csv")
```


<!-- The user Bobby can also be removed again. -->

```{r rmbobby, include=FALSE}
# authenticate
iauth("rods", "rods")

# remove user bobby
rirods:::iadmin(action = "remove", target = "user", arg2 = "bobby")

# check if bobby is removed
ils()

# clean-up irods file
file.remove("rirods.irods")
```

