---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rirods2

<!-- badges: start -->
<!-- badges: end -->

The goal of rirods2 is a pure R client for iRODS

## Installation

You can install the development version of rirods2 like so:

``` r
# install.packages("devtools")
devtools::install_github("MartinSchobben/rirods2")
```

## Example

This is a basic example which shows you how to quickly launch a local iRODS server. For more information check https://github.com/irods/irods_demo.

```{bash, eval=FALSE}
# clone the repo
# git clone https://github.com/irods/irods_demo
# initiate git submodules
# git submodule update --init
# to start
cd ../irods_demo
docker-compose up
```

Suppose Bobby would like to upload an R object from his current R session to an iRODS collection. For this, use the iput command:

```{r put}
library(rirods2)

# login as bobby
iauth("bobby", "passWORD")

# some data
foo <- data.frame(x = c(1, 8, 9), y = c("x", "y", "z"))

# check where we are
ipwd()

# store
iput(foo)

# check if file is stored
ils()
```

If Bobby wanted to copy the foo R object from an iRODS collection to his local directory, he would use
iget:

```{r get}
# retrieve in native R format
iget("foo")

# delete object
irm("foo", trash = FALSE)

# check if object is removed
ils()
```

Possibly Bobby does not want a native R object to be stored on iRODS but a file type that can be accessed by other programs.

```{r filetypes}
library(readr)

# creates a csv file of foo
write_csv(foo, "foo.csv")

# send file
iput("foo.csv")

# check whether it is stored
ils()
```

```{r rmfile1, include= FALSE}
file.remove("foo.csv")
```

Later on somebody else might want to download this file again and store it locally.

```{r csv}
# retrieve it again later
iget("foo.csv")
read_csv("foo.csv")

# delete object
irm("foo.csv", trash = FALSE)
```

```{r rmfile2, include= FALSE}
file.remove("foo.csv")
```



<!-- The user Bobby can also be removed again. -->

```{r rmbobby, include=FALSE}
# authenticate
iauth()

# remove user bobby
iadmin(action = "remove", target = "user", arg2 = "bobby")

# check if bobby is removed
ils()
```

## Stop your local iRODS server

```{bash, eval=FALSE}
docker-compose down
```
