---
title: "Developing rirods"
output: 
  rmarkdown::html_vignette:
    md_extensions: [ 
      "-autolink_bare_uris" 
    ]
vignette: >
  %\VignetteIndexEntry{Developing rirods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Contributing to rirods development

To contribute to rirods fork the source files from: 
https://github.com/irods/irods_client_library_rirods, and make a Pull Request. 

## Prerequisites R

The following R packages are required for development: `jsonlite`, `readr`, 
`testthat` (>= 3.0.0) and `withr`, and `httr2` and `httptest2`. For convenience,
one can also use `renv` () and `renv::restore()` to install all 
dependencies. The `usethis` and the `devtools` packages are further 
suggestions to streamline development. See https://devtools.r-lib.org/ and 
https://usethis.r-lib.org/ for common development tasks and setups covered by 
these packages.

Development of `rirods` relies on the `httptest2` which enables testing of the 
logic on the R side of the API without requiring access to the remote service.
It does this by recording the API response as mock files. These mock files are 
stored in several directories under `.tests/testthat/`. To enable testing
against the real API one needs to remove these mock files. The helper 
`remove_mock_files()` simplifies this task. This function is not exported and 
needs to be sourced (e.g., using `devtools::load_all()`) and subsequently 
called.

## Prerequisites iRODS

The unit test included in the source files can either be run against a local or 
remote iRODS C++ REST API service, version 0.9.3 or later - 
https://github.com/irods/irods_client_rest_cpp.

### Local server

Launch a local demonstration iRODS service (including the REST API):

```{bash, eval=FALSE}
# clone the repository
git clone --recursive https://github.com/irods/irods_demo
# start the REST API
cd irods_demo
docker-compose up -d nginx-reverse-proxy
```

This will result in the demonstration REST API running at http://localhost/irods-rest/0.9.3 (or later version). 
No further configuration in R is needed for development (except for removing 
possible mock files). For example, to run the unit tests against this local 
server use: `devtools::test()`.

For miscellaneous development tasks one can, furthermore, use the non-exported 
function: `local_create_irods()`. This test fixture is almost the same as the
exported function `create_irods()` except that it uses a temporary directory as 
the working directory thereby preventing cluttering of the package source files.
When restarting the R session (possibly with `CTRL + SHIFT + F10`) or running 
`withr::deferred_run()` the working directory is restored to the 
package directory. 

### Remote server

Testing against a remote server requires the specification of server 
information. This information needs to be entered in the `.Rprofile` file:

```{r prof, eval=FALSE}
# dependency management
options(renv.settings.snapshot.type = "explicit")
source("renv/activate.R")

# development key (create key with httr2::secret_make_key() and place in user
# level environment variables. One can use usethis::edit_r_environ() for this.
# Store the key under "DEV_KEY_IRODS")

# irods environment variables for development
Sys.setenv(DEV_HOST_IRODS = "nLO8T0IpHCT2kXklYE-IB0HjYpNkg5wN4ZKk7TPSvkwGf9FauDR-O5mVI-mmD2_zNFGLoaVcAgYl")
Sys.setenv(DEV_ZONE_PATH_IRODS = "gunp5Ts08NHfpKAdbFCBuilVb7Wjnxojbt7bA1e-9Q")
Sys.setenv(DEV_USER = "ZGlORquE2G6BIPS5JAcuPcngmBB6Wg")
Sys.setenv(DEV_PASS = "ZGlORquE2G6BIPS5JAcuPcngmBB6Wg")
```

The server information is stored in the environmental variables `DEV_HOST_IRODS`
,`DEV_ZONE_PATH_IRODS`, `DEV_USER`, and `DEV_PASS`. Server information is 
provided in scrambled format to prevent leakage of sensitive information to 
GitHub. 

The following step are needed to scramble this information.

1) Generate package development key:

```{r key, eval=FALSE}
httr2::secret_make_key()
```

2) Store the package development key as an environmental variable at the user 
level. For convenience one can use `usethis::edit_r_environ()` which opens the 
`.Renviron` file. Enter your key by replacing `<key>` as follows:

```
DEV_KEY_IRODS="<key>"
```

3) Scramble the iRODS server information with `httr2::secret_encrypt()`. 
For example, scramble the host name as follows:

```{r host, eval=FALSE}
httr2::secret_encrypt("http://myserver/irods-rest/0.9.3" ,"DEV_KEY_IRODS")
```

4) Store the scrambled information at the designated places in the project level 
`.Rprofile`: host name under `DEV_HOST_IRODS`, zone path under 
`DEV_ZONE_PATH_IRODS` (e.g., `"/tempZone/home"`), user name under `DEV_USER`, 
and password under `DEV_PASS`.

5) Restart the R sessions (possibly with `CTRL + SHIFT + F10`).

Note that the unit tests are built against a clean server with only one user 
(i.e., `DEV_USER`). In addition, one should not forget to remove the mock files
(e.g., with `remove_mock_files()`, see above) to test against the real server.

