---
output: 
  github_document:
     md_extensions: [ 
      "-autolink_bare_uris" 
    ]
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
if (!requireNamespace("readr", quietly = TRUE)) install.packages("readr")
if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")

here::i_am("inst/bench/README.Rmd")
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)

make_table <- function(type = "files", cap = "**") {
  read_tsv(paste0(c("main-", "dev-"), type, ".tsv"), show_col_types = FALSE) |> 
    relocate(src, ref, `file size` = size, .before = expression) |> 
    select(
      !c(result, memory, time, `gc/sec`,	n_itr,	n_gc, gc, 
         total_time, min, `itr/sec`), 
      `memory allocated` = mem_alloc) |> 
    knitr::kable(digits = 2, align = "r", format = "html", caption = cap)
}

make_plot <- function(type = "files") {
  read_tsv(paste0(c("main-", "dev-"), type, ".tsv"), show_col_types = FALSE) |> 
    mutate(
      `memory allocated (Mb)` = 
        case_when(
          endsWith(mem_alloc, "GB") ~ parse_number(mem_alloc) * 1e3,
          endsWith(mem_alloc, "MB") ~ parse_number(mem_alloc),
          endsWith(mem_alloc, "KB") ~ parse_number(mem_alloc) / 1e3
        ),
      `median time (sec)` = 1 / `itr/sec`
      ) |> 
    pivot_longer(c(`median time (sec)`, `memory allocated (Mb)`)) |> 
    ggplot(aes(x = size / 1e6, y = value, color = paste0(src, "@", ref))) +
      geom_point() +
      geom_line() +
      scale_x_log10() +
      labs(x = "Size (Mb)", y = "", color = "") + 
      theme(legend.position = "top", legend.box = "vertical") +
      guides(color = guide_legend(nrow = 3, byrow = TRUE)) +
      facet_grid(cols = vars(expression), rows = vars(name), scales = "free")
  
}
```

# rirods benchmarks

## Prerequisites

For the benchmarks one needs to connect to the iRODS C++ REST/HTTP API.

Launch a local demonstration iRODS service (including the REST/HTTP API):

```{bash, eval=FALSE}
# clone the repository
git clone --recursive https://github.com/irods/irods_demo
# start the REST API
cd irods_demo
docker compose up -d nginx-reverse-proxy
```

This will result in the demonstration REST/HTTP API running at http://localhost/irods-rest/0.9.3 or http://localhost:9001/irods-http-api/0.2.0.

## Usage benchmarks

The benchmarks are based on the R package *bench*. The two parameters of interest are `median`, which is the sample median time and `mem_alloc`, which is the total amount of memory allocated by R while running the expression.

To perform a benchmark use GNU make:

```{bash, eval=FALSE}
make
```

## Results

The results are based on the HEAD of the main branch on the iRODS organization repository and the local package. The test case uses `iput()` and `iget()` with a `count` smaller than the file size.

## Performance iget and iput

```{r table1, echo=FALSE}
make_table(cap = "<i>Table 1: Comparing allocation of memory and runtime for `iget()` and `iput()` for different file sizes.</i>")
```

```{r plot1, echo=FALSE, fig.cap="<i>Fig 1: Comparing allocation of memory and runtime for `iget()` and `iput()` for different file sizes.</i>"}
make_plot()
```


## Performance isaveRDS and ireadRDS

```{r table2, echo=FALSE}
make_table(type = "objects", cap = "<i>Table 2: Comparing allocation of memory and runtime for `isaveRDS()` and `ireadRDS()` for different object sizes.</i>")
```

```{r plot2, echo=FALSE, fig.cap="<i>Fig 2: Comparing allocation of memory and runtime for `isaveRDS()` and `ireadRDS()` for different object sizes.</i>"}
make_plot(type = "objects")
```
